Tufts BAT - R data visualization workshop
By Kyle Krell

```{r}

#install required libaries
install.packages("dplyr")
install.packages("ggplot2")
install.packages("ggthemes")
install.packages("ggimage")
install.packages("tidyr")
install.packages("baseballr")
```

```{r}

#load libraries
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggimage)
library(tidyr)
library(baseballr)
```

```{r}
#Viewing certain functions in the BaseballR package. There are so many to choose from!
baseballr::fg_park(2023)
baseballr::sptrc_league_payrolls(2024)
baseballr::mlb_draft(2023)
```

```{r}
#view the names of the columns in our data 
#colnames(baseballr::fg_batter_leaders())

#save data to a variable called 'data' so we can call it in the future
#This data contains all individual batter seasons from 2023
data <- baseballr::fg_batter_leaders()
```

DATA MANIPULATON
```{r}
#get batters with over 150 plate appearances
filter(data, PA > 150)
```
Notice the size of our data has decreased to 404, meaning we only have 404 people 
with over 150 PA.

Now, we will do the same thing using the pipe operator.
```{r}
data |> filter(PA > 150)
```
We get the same result, and using the pipe operator, we can make our code more 
readable.

```{r}
#save filtered dataset to new variable
filtered_data <- data |> filter(PA > 150)
```

```{r}
#generate scatter plot where x is barrel percentage and y is hard hit pct
#the '+ geom_point' means that we get a scatterplot
#aes means aesthetic here, and it allows us to define our x and y axes

filtered_data |> ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + geom_point()

#can also break this up into multiple lines
filtered_data |>
  ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + 
  geom_point()
```
```{r}
filtered_data |>
  ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + 
  geom_point(color = 'magenta') #can change point color with color argument

filtered_data |>
  ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + 
  geom_point(color = 'magenta') + #can change point color with color argument 
  geom_smooth(method = 'lm', color = 'black', se = FALSE) #this function adds a trendline, se = FALSE deletes error bar


#flipping the order makes the points appear in front of the trendline
filtered_data |>
  ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + 
  geom_smooth(method = 'lm', color = 'black', se = FALSE) + 
  geom_point(color = 'magenta')  

#using the theme function
filtered_data |>
  ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + 
  geom_point(color = 'magenta') +
  geom_smooth(method = 'lm', color = 'black', se = FALSE) +
  theme_clean() #this theme function gives your plot a theme, can help coloring, font etc 
  
```
Now, let's add labels to our graph.
```{r}
filtered_data |>
  ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + 
  geom_point(color = 'magenta') +
  geom_smooth(method = 'lm', color = 'black', se = FALSE) +
  #add axis labels and title
  xlab('Barrel Percent') +
  ylab('Hard Hit Percent') +
  labs(title = 'Barrel vs. Hard Hit PCT' ) +
  theme_clean() 

#center the title, but we have to do it after the theme as our centering has to override the theme's title
filtered_data |>
  ggplot(aes(x = Barrel_pct, y = HardHit_pct)) + 
  geom_point(color = 'red3') +
  geom_smooth(method = 'lm', color = 'black', se = FALSE) +
  #add axis labels and title
  xlab('Barrel Percent') + ylab('Hard Hit Percent') +
  labs(title = 'Barrel % vs. Hard Hit %') +
  theme_clean() +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 14, face = 'bold')) +
  ylim(0, NA)


#save plot
ggsave('barrel_vs_hardhit.png', height = 10, width = 14, dpi = 'retina')
```
Now, let's create a faceplot!
```{r}
#load pitching data
pitchdat <- baseballr::fg_pitch_leaders()

pitchdat <- pitchdat |>
  filter(TBF > 60)

```

```{r}
#create basic scatterplot
pitchdat |>
  ggplot(aes(x = Swing_pct, y = Contact_pct)) + 
  geom_point(color = 'blue') +
  geom_smooth(method = 'lm', color = 'red', se = FALSE) +
  #add axis labels and title
  xlab('Swing percentage') +
  ylab('Contact percentage') +
  labs(title = 'Swing% vs. Contact%' ) +
  theme_clean() 
```
# Manipulating data for face plot
```{r}
# URL formula for player faces
# https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current.png/w_426,q_auto:best/v1/people/554430/headshot/67/current
# The formula is "https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current.png/w_426,q_auto:best/v1/people/" + [the player's MLB ID] + "/headshot/67/current"

pitchdat$xMLBAMID

#create URL column that will allow us to pull player faces

paste0('https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current.png/w_426,q_auto:best/v1/people/',
       pitchdat$xMLBAMID,
       '/headshot/67/current')

pitchdat <- pitchdat |> 
  mutate(URLs = 
           paste0('https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current.png/w_426,q_auto:best/v1/people/',
                  pitchdat$xMLBAMID,
                  '/headshot/67/current'))

#pitchdat$URLs
```

# Making the face plot
```{r}
# Scatter plot from before
pitchdat |> 
  ggplot(aes(x = pfx_Swing_pct, y = pfx_Contact_pct)) + 
  geom_point(color = 'red3') +
  geom_smooth(method = 'lm', color = 'black', se = FALSE) + 
  xlab('Swing Percent') + ylab('Contact Percent') +
  labs(title = 'Swing % vs. Contact %') +
  theme_clean() +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 14, face = 'bold'))

# Changing geom_point to geom_image
pitchdat |> 
  ggplot(aes(x = pfx_Swing_pct, y = pfx_Contact_pct)) + 
  geom_image(image = pitchdat$URLs, asp = 16/9, size = 0.03) +
  xlab('Swing Percent') + ylab('Contact Percent') +
  labs(title = 'Swing % vs. Contact %') +
  theme_clean() +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 14, face = 'bold'))
```

# Small aesthetic changes to face plot
```{r}
# Adding median dashed lines
pitchdat |> 
  ggplot(aes(x = pfx_Swing_pct, y = pfx_Contact_pct)) + 
  geom_hline(yintercept = median(pitchdat$pfx_Contact_pct), linetype = 'dashed') +
  geom_vline(xintercept = median(pitchdat$pfx_Swing_pct), linetype = 'dashed') +
  geom_image(image = pitchdat$URLs, asp = 16/9, size = 0.03) +
  xlab('Swing Percent') + ylab('Contact Percent') +
  labs(title = 'Swing % vs. Contact %') +
  theme_clean() +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 14, face = 'bold'))

# Scaling axes
pitchdat |> 
  ggplot(aes(x = pfx_Swing_pct * 100, y = pfx_Contact_pct * 100)) + 
  geom_hline(yintercept = median(pitchdat$pfx_Contact_pct) * 100, linetype = 'dashed') +
  geom_vline(xintercept = median(pitchdat$pfx_Swing_pct) * 100, linetype = 'dashed') +
  geom_image(image = pitchdat$URLs, asp = 16/9, size = 0.03) +
  xlab('Swing Percent') + ylab('Contact Percent') +
  labs(title = 'Swing % vs. Contact %') +
  theme_clean() +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 14, face = 'bold'))

# Flipping y-axis
pitchdat |> 
  ggplot(aes(x = pfx_Swing_pct * 100, y = pfx_Contact_pct * 100)) + 
  geom_hline(yintercept = median(pitchdat$pfx_Contact_pct) * 100, linetype = 'dashed') +
  geom_vline(xintercept = median(pitchdat$pfx_Swing_pct) * 100, linetype = 'dashed') +
  geom_image(image = pitchdat$URLs, asp = 16/9, size = 0.03) +
  xlab('Swing Percent') + ylab('Contact Percent') +
  labs(title = 'Swing % vs. Contact %') +
  theme_clean() +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 14, face = 'bold')) + 
  scale_y_reverse()

# Saving image
ggsave('swing_vs_contact.png', dpi = 'retina', height = 10, width = 14)
```

# Summarizing code
```{r}
pitchdat <- fg_pitch_leaders()

pitchdat <- pitchdat |> 
  filter(TBF > 300)

pitchdat <- pitchdat |> 
  mutate(URLs = 
           paste0('https://img.mlbstatic.com/mlb-photos/image/upload/d_people:generic:headshot:67:current.png/w_426,q_auto:best/v1/people/',
                  pitchdat$xMLBAMID,
                  '/headshot/67/current'))

pitchdat |> 
  ggplot(aes(x = pfx_Swing_pct * 100, y = pfx_Contact_pct * 100)) + 
  geom_hline(yintercept = median(pitchdat$pfx_Contact_pct) * 100, linetype = 'dashed') +
  geom_vline(xintercept = median(pitchdat$pfx_Swing_pct) * 100, linetype = 'dashed') +
  geom_image(image = pitchdat$URLs, asp = 16/9, size = 0.03) +
  xlab('Swing Percent') + ylab('Contact Percent') +
  labs(title = 'Swing % vs. Contact %') +
  theme_clean() +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 14, face = 'bold')) + 
  scale_y_reverse()

ggsave('swing_vs_contact2.png', dpi = 'retina', height = 10, width = 14)
```






# How to get season play-by-play data
```{r}
game_pks <- mlb_schedule(2023, level_ids = 1) |>
  filter(game_type %in% c('D', 'F', 'L', 'R', 'W')) |>
  mutate(game_pks = as.list(game_pk)) |>
  pull(game_pks)

season_pbp <- list()
count = 0
for (game_pk in game_pks) {
  result <- tryCatch({
    game_pbp <- mlb_pbp(game_pk = game_pk)
    game_pbp},
    
    error = function(err) {
      message(paste('Error occurred for game_pk', game_pk))
      NULL})
  
  if (!is.null(result)) {
    season_pbp[[length(season_pbp) + 1]] <- result}
  
  count = count + 1
  message(paste('Games complete:', count))}

fill_tibble <- function(df, cols) {
  missing_cols <- setdiff(cols, colnames(df))
  df[, missing_cols] <- NA
  return(df)}

all_cols <- season_pbp |>
  map(names) |>
  reduce(union)

pbp_filled <- season_pbp |>
  map(~ fill_tibble(.x, all_cols))

season_pbp <- bind_rows(pbp_filled)
```

# Loading in resulting season play-by-play
```{r}
season_pbp <- read.csv('/Users/kylekrell/Desktop/Learning_to_Code/mlb_pbp.csv')
```

# Merging with another dataset
```{r}
season_pbp <- left_join(season_pbp, 
          mlb_schedule(2023, level_ids = 1) |> filter(game_type %in% c('D', 'F', 'L', 'R', 'W')) |> select(-game_date),
          by = 'game_pk')
```

# Various data manipulation
```{r}
stadium_ks <- season_pbp |> 
  group_by(venue_name, game_pk, atBatIndex) |> 
  summarise(K = ifelse(result.eventType == 'strikeout', T, F)) |> 
  slice(1) |>
  group_by(venue_name) |> 
  summarise(K = sum(ifelse(K == T, 1, 0))) |> 
  filter(!venue_name %in% c('Estadio Alfredo Harp Helu', 'London Stadium', 'Muncy Bank Ballpark'))

stadium_ks$team <- c('Brewers', 'Angels', 'Cardinals', 'Diamondbacks', 'Mets', 'Phillies', 'Tigers', 'Rockies', 'Dodgers', 'Red Sox',
                     'Rangers', 'Reds', 'White Sox', 'Royals', 'Astros', 'Nationals', 'Athletics', 'Giants', 'Orioles', 'Pirates', 
                     'Padres', 'Guardians', 'Blue Jays', 'Mariners', 'Twins', 'Rays', 'Braves', 'Cubs', 'Yankees', 'Marlins')
  
stadium_ks <- stadium_ks |> 
  left_join(fg_park(2023), by = c('team' = 'home_team'))
  
stadium_ks <- mlb_teams(season = 2023) |> 
  filter(sport_name == 'Major League Baseball') |> 
  select(team_full_name, team_name, team_abbreviation) |> 
  mutate(team_name = ifelse(team_name == 'D-backs', 'Diamondbacks', team_name)) |> 
  mutate(team_abbreviation = case_when(
    team_abbreviation == 'AZ' ~ 'ARI',
    team_abbreviation == 'KC' ~ 'KCR',
    team_abbreviation == 'SD' ~ 'SDP',
    team_abbreviation == 'SF' ~ 'SFG',
    team_abbreviation == 'TB' ~ 'TBR',
    team_abbreviation == 'WSH' ~ 'WSN',
    team_abbreviation == 'CWS' ~ 'CHW',
    TRUE ~ team_abbreviation),
    URLs = paste0('https://cdn.ssref.net/req/202312151/tlogo/br/',
                  team_abbreviation, '-2023.png')) |>
  right_join(stadium_ks, by = c('team_name' = 'team'))
```

# Making the plot
```{r}
stadium_ks |> 
  ggplot(aes(x = K, y = jitter(so)))+
  theme_bw()+
  geom_hline(yintercept = mean(stadium_ks$so), lty = 'dashed')+
  geom_vline(xintercept = mean(stadium_ks$K), lty = 'dashed')+
  geom_image(image = stadium_ks$URLs, asp = 16/9, size = .025)+
  labs(title = 'Stadium Strikeouts and SO Park Factor')+
  theme(axis.title = element_text(size=14))+
  xlab('Stadium Strikeouts')+
  ylab('SO Park Factor')+
  theme(panel.grid.minor = element_blank())+
  theme(axis.text = element_text(size = 10),
        plot.title = element_text(size = 16, hjust = 0.5, face = 'bold'))+
  scale_y_continuous(breaks = scales::pretty_breaks(n=8))+
  scale_x_continuous(breaks = scales::pretty_breaks(n=8))
```

# More manipulation
```{r}
away_ks <- season_pbp |> 
  group_by(away_team, game_pk, atBatIndex) |> 
  summarise(TeamKs = ifelse(result.eventType == 'strikeout', T, F)) |> 
  slice(1) |> 
  group_by(away_team) |> 
  summarise(TeamKs = sum(ifelse(TeamKs == T, 1, 0)))

away_ks <- baseballr::mlb_teams(season = 2023) |> 
  filter(sport_name == 'Major League Baseball') |> 
  select(team_full_name, team_name, team_abbreviation) |> 
  mutate(team_name = ifelse(team_name == 'D-backs', 'Diamondbacks', team_name)) |> 
  mutate(team_abbreviation = case_when(
    team_abbreviation == 'AZ' ~ 'ARI',
    team_abbreviation == 'KC' ~ 'KCR',
    team_abbreviation == 'SD' ~ 'SDP',
    team_abbreviation == 'SF' ~ 'SFG',
    team_abbreviation == 'TB' ~ 'TBR',
    team_abbreviation == 'WSH' ~ 'WSN',
    team_abbreviation == 'CWS' ~ 'CHW',
    TRUE ~ team_abbreviation),
    URLs = paste0('https://cdn.ssref.net/req/202312151/tlogo/br/',
                  team_abbreviation, '-2023.png')) |>
  right_join(away_ks, by = c('team_full_name' = 'away_team'))

away_ks <- away_ks |> 
  left_join(fg_park(2023), by = c('team_name' = 'home_team'))
```

# More plots
```{r}
away_ks |> 
  ggplot(aes(x = TeamKs, y = jitter(so)))+
  theme_bw()+
  geom_hline(yintercept = mean(away_ks$so), lty = 'dashed')+
  geom_vline(xintercept = mean(away_ks$TeamKs), lty = 'dashed')+
  geom_image(image = stadium_ks$URLs, asp = 16/9, size = .025)+
  labs(title = 'Team Away Strikeouts and SO Park Factor')+
  theme(axis.title = element_text(size=14))+
  xlab('Team Away Strikeouts (Pitching and Hitting)')+
  ylab('SO Park Factor')+
  theme(panel.grid.minor = element_blank())+
  theme(axis.text = element_text(size = 10),
        plot.title = element_text(size = 16, hjust = 0.5, face = 'bold'))+
  scale_y_continuous(breaks = scales::pretty_breaks(n=8))+
  scale_x_continuous(breaks = scales::pretty_breaks(n=8))
```
